Practical Open-Loop Optimistic Planning

### 2019-09-14 paper/main.tex

\emph{Monte-Carlo Tree Search} (\texttt{MCTS}) algorithms were historically motivated by the application of computer Go, and made a first appearance in the CrazyStone software \citet{Coulom2006}. They were later reformulated in the setting of Multi-Armed Bandits by \citet{Kocsis2006} with their \emph{Upper Confidence bounds applied to Trees} (\texttt{UCT}) algorithm. Despite its popularity \citep{Silver2016,Silver2017,Silver2018}, \texttt{UCT} has been shown to suffer from several limitations: its sample complexity can be at least doubly-exponential for some problems (e.g. when a narrow optimal path is hidden in a suboptimal branch), which is much worse than uniform planning \citep{Coquelin2007}. The \texttt{Sparse Sampling} algorithm of \citet{Kearns2002} achieves better worst-case performance, but it is still non-polynomial and doesn't adapt to the structure of the MDP. In stark contrast, the \emph{Optimistic Planning for Deterministic systems} (\texttt{OPD}) algorithm considered by \citet{Hren2008} in the case of deterministic transitions and rewards exploits the structure of the cumulative discounted reward to achieve a problem-dependent polynomial bound on sample complexity. A similar line of work in a deterministic setting is that of \texttt{SOOP} and \texttt{OPC} by \cite{Busoniu2013,Busoniu2018} though they focus on continuous action spaces. \texttt{OPD} was later extended to stochastic systems with the \emph{Open-Loop Optimistic Planning} (\OLOP) algorithm introduced by \citet{Bubeck2010} in the open-loop setting: we only consider sequences of actions independently of the states that they lead to. This restriction in the space of policies causes a loss of optimality, but greatly simplifies the planning problem in the cases where the state space is large or infinite. More recent work such as \texttt{St0p} \citep{Szorenyi2014} and \texttt{TrailBlazer} \citep{Grill2016} focus on the probably approximately correct (PAC) framework: rather than simply recommending an action to maximise the expected rewards, they return an $\epsilon$-approximation of the value at the root that holds with high probability. This highly demanding framework puts a severe strain on these algorithms that were developed for theoretical analysis only and cannot be applied to real problems.

@@

We have performed some numerical experiments to evaluate and compare the following planning algorithms\footnote[1]{The source code is available at \url{https://eleurent.github.io/kl-olop/}}:
\begin{itemize}
    \item \texttt{Random}: returns a random action, we use it as a minimal performance baseline.
    \item \texttt{OPD}: the \emph{Optimistic Planning for Deterministic systems} from \citep{Hren2008}, used as a baseline of optimal performance. This planner is only suited for deterministic environments, and exploits this property to obtain faster rates. However, it is expected to fail in stochastic environments.
    \item \OLOP: as described in section \ref{sec:kl-olop-olop}.\footnote[2]{Note that we use the lazy version of $\OLOP$ and $\KLOLOP$ presented in Section \ref{sec:time-complexity}, otherwise the exponential running-time would have been prohibitive.}
    \item \KLOLOP: as described in section \ref{sec:kl-olop-kl-olop}.\footnotemark[2]
    \item \texttt{KL-OLOP}(1): an aggressive version of \KLOLOP where we used $f_1(m) = \log M$ instead of $f_2(m)$. This threshold function makes the upper bounds even tighter, at the cost of an increased probability of violation. Hence, we expect this solution to be more efficient in close-to-deterministic environments. However, since we have no theoretical guarantee concerning its regret as we do with $\KLOLOP$, it might not be conservative enough and converge too early to a suboptimal sequence, especially in highly stochastic environments.
\end{itemize}

@@

	\includegraphics[width=0.64\textwidth]{img/tree-ODP_svg-tex} \includegraphics[width=0.64\textwidth]{img/tree-OLOP_svg-tex}
	\includegraphics[width=0.64\textwidth]{img/tree-KL-OLOP_svg-tex}

@@

The results of our experiments are shown in Figure \ref{fig:experiments}. The \texttt{ODP} algorithm converges very quickly to the optimal return in the two first environments, shown in Figure \ref{sub:highway} and Figure \ref{sub:gridworld}, because it exploits their deterministic nature: it needs neither to estimate the rewards through upper-confidence bounds nor to sample whole sequences all the way from the root when expanding a leaf, which provides a significant speedup. It can be seen as an oracle allowing to measure the conservativeness of stochastic planning algorithms. And indeed, even before introducing stochasticity, we can see that \OLOP performs quite badly on the two environments, only managing to solve them with a budget in the order of $10^{3.5}$. In stark contrast, \KLOLOP makes a much better use of its samples and reaches the same performance an order of magnitude faster. This is illustrated by the expanded trees shown in Figure \ref{fig:trees}: \ODP exploits the deterministic setting and produces a sparse tree densely concentrated around the optimal trajectory. Conversely, the tree developed by \OLOP is evenly balanced, which suggests that \OLOP behaves as uniform planning as hypothesised in Section \ref{sec:kl-olop-behaviour}. \KLOLOP is more efficient and expands a highly unbalanced tree, exploring the same regions as \ODP. Furthermore, in the stochastic gridworld environment shown in Figure \ref{sub:gridworld_stoch}, we observe that the deterministic \texttt{ODP} planner's performance saturates as it settles to suboptimal trajectories, as expected. Conversely, the stochastic planners all find better-performing open-loop policies, which justifies the need for this framework. Again, \KLOLOP converges an order of magnitude faster than \OLOP. Finally, \texttt{KL-OLOP}(1) enjoys good performance overall and displays the most satisfying trade-off between aggressiveness in deterministic environments and conservativeness in stochastic environments; hence we recommend this tuning for practical use.

@@

\bibliographystyle{splncs04}
\bibliography{references}

@@

%\section{Examples of expanded trees}
%
%The trees expanded by different planning algorithms in the Highway environment are displayed in Figure \ref{fig:trees}. They were all generated in the same initial root state, under the same budget of $n=10^3$ calls to the generative model. We observe that \ODP exploits the deterministic setting efficiently and produces a very sparse tree densely concentrated around the optimal trajectory. Conversely, the tree developed by the \OLOP algorithm is quite evenly balanced, which suggests that \OLOP behaves as uniform planning as hypothesised in \ref{sec:kl-olop-behaviour}. In stark contrast, \KLOLOP is much more efficient and produced a highly unbalanced tree, exploring the same regions of the tree as \ODP. 
%
%\begin{figure}[pht]
%    \centering
%
%    \includegraphics[width=0.64\textwidth]{img/tree-ODP_svg-tex}    \includegraphics[width=0.64\textwidth]{img/tree-OLOP_svg-tex}
%    \includegraphics[width=0.64\textwidth]{img/tree-KL-OLOP_svg-tex}
%
%    \caption{The look-ahead trees expanded by different planning algorithms for a budget $n=10^3$, shown down to depth 6. The width of edges is proportional to the nodes visit count $T_a(M)$.}
%    \label{fig:trees}
%\end{figure}

>>>>

\emph{Monte-Carlo Tree Search} (\texttt{MCTS}) algorithms were historically motivated by the application of computer Go, and made a first appearance in the CrazyStone software \citet{Coulom2006}. They were later reformulated in the setting of Multi-Armed Bandits by \citet{Kocsis2006} with their \emph{Upper Confidence bounds applied to Trees} (\texttt{UCT}) algorithm. Despite its popularity \citep{Silver2016,Silver2017,Silver2018}, \texttt{UCT} has been shown to suffer from several limitations: its sample complexity can be at least doubly-exponential for some problems (e.g. when a narrow optimal path is hidden in a suboptimal branch), which is much worse than uniform planning \citep{Coquelin2007}. The \texttt{Sparse Sampling} algorithm of \citet{Kearns2002} achieves better worst-case performance, but it is still non-polynomial and doesn't adapt to the structure of the MDP. In stark contrast, the \emph{Optimistic Planning for Deterministic systems} (\OPD) algorithm considered by \citet{Hren2008} in the case of deterministic transitions and rewards exploits the structure of the cumulative discounted reward to achieve a problem-dependent polynomial bound on sample complexity. A similar line of work in a deterministic setting is that of \texttt{SOOP} and \texttt{OPC} by \cite{Busoniu2013,Busoniu2018} though they focus on continuous action spaces. \OPD was later extended to stochastic systems with the \emph{Open-Loop Optimistic Planning} (\OLOP) algorithm introduced by \citet{Bubeck2010} in the open-loop setting: we only consider sequences of actions independently of the states that they lead to. This restriction in the space of policies causes a loss of optimality, but greatly simplifies the planning problem in the cases where the state space is large or infinite. More recent work such as \texttt{St0p} \citep{Szorenyi2014} and \texttt{TrailBlazer} \citep{Grill2016} focus on the probably approximately correct (PAC) framework: rather than simply recommending an action to maximise the expected rewards, they return an $\epsilon$-approximation of the value at the root that holds with high probability. This highly demanding framework puts a severe strain on these algorithms that were developed for theoretical analysis only and cannot be applied to real problems.

@@

We have performed some numerical experiments to evaluate and compare the following planning algorithms\footnote[1]{The source code is available at \url{https://eleurent.github.io/kl-olop/}}:
\begin{itemize}
    \item \texttt{Random}: returns a random action, we use it as a minimal performance baseline.
    \item \OPD: the \emph{Optimistic Planning for Deterministic systems} from \citep{Hren2008}, used as a baseline of optimal performance. This planner is only suited for deterministic environments, and exploits this property to obtain faster rates. However, it is expected to fail in stochastic environments.
    \item \OLOP: as described in section \ref{sec:kl-olop-olop}.\footnote[2]{Note that we use the lazy version of $\OLOP$ and $\KLOLOP$ presented in Section \ref{sec:time-complexity}, otherwise the exponential running-time would have been prohibitive.}
    \item \KLOLOP: as described in section \ref{sec:kl-olop-kl-olop}.\footnotemark[2]
    \item \texttt{KL-OLOP}(1): an aggressive version of \KLOLOP where we used $f_1(m) = \log M$ instead of $f_2(m)$. This threshold function makes the upper bounds even tighter, at the cost of an increased probability of violation. Hence, we expect this solution to be more efficient in close-to-deterministic environments. However, since we have no theoretical guarantee concerning its regret as we do with $\KLOLOP$, it might not be conservative enough and converge too early to a suboptimal sequence, especially in highly stochastic environments.
\end{itemize}

@@

	\includegraphics[width=0.64\textwidth]{img/tree_OPD_svg-tex} \includegraphics[width=0.64\textwidth]{img/tree_OLOP_svg-tex}
	\includegraphics[width=0.64\textwidth]{img/tree_KL-OLOP_svg-tex}

@@

The results of our experiments are shown in Figure \ref{fig:experiments}. The \OPD algorithm converges very quickly to the optimal return in the two first environments, shown in Figure \ref{sub:highway} and Figure \ref{sub:gridworld}, because it exploits their deterministic nature: it needs neither to estimate the rewards through upper-confidence bounds nor to sample whole sequences all the way from the root when expanding a leaf, which provides a significant speedup. It can be seen as an oracle allowing to measure the conservativeness of stochastic planning algorithms. And indeed, even before introducing stochasticity, we can see that \OLOP performs quite badly on the two environments, only managing to solve them with a budget in the order of $10^{3.5}$. In stark contrast, \KLOLOP makes a much better use of its samples and reaches the same performance an order of magnitude faster. This is illustrated by the expanded trees shown in Figure \ref{fig:trees}: \OPD exploits the deterministic setting and produces a sparse tree densely concentrated around the optimal trajectory. Conversely, the tree developed by \OLOP is evenly balanced, which suggests that \OLOP behaves as uniform planning as hypothesised in Section \ref{sec:kl-olop-behaviour}. \KLOLOP is more efficient and expands a highly unbalanced tree, exploring the same regions as \OPD. Furthermore, in the stochastic gridworld environment shown in Figure \ref{sub:gridworld_stoch}, we observe that the deterministic \OPD planner's performance saturates as it settles to suboptimal trajectories, as expected. Conversely, the stochastic planners all find better-performing open-loop policies, which justifies the need for this framework. Again, \KLOLOP converges an order of magnitude faster than \OLOP. Finally, \texttt{KL-OLOP}(1) enjoys good performance overall and displays the most satisfying trade-off between aggressiveness in deterministic environments and conservativeness in stochastic environments; hence we recommend this tuning for practical use.

@@

\bibliographystyle{splncs04}
\bibliography{biblio}

@@

%\section{Examples of expanded trees}
%
%The trees expanded by different planning algorithms in the Highway environment are displayed in Figure \ref{fig:trees}. They were all generated in the same initial root state, under the same budget of $n=10^3$ calls to the generative model. We observe that \OPD exploits the deterministic setting efficiently and produces a very sparse tree densely concentrated around the optimal trajectory. Conversely, the tree developed by the \OLOP algorithm is quite evenly balanced, which suggests that \OLOP behaves as uniform planning as hypothesised in \ref{sec:kl-olop-behaviour}. In stark contrast, \KLOLOP is much more efficient and produced a highly unbalanced tree, exploring the same regions of the tree as \OPD. 
%
%\begin{figure}[pht]
%    \centering
%
%    \includegraphics[width=0.64\textwidth]{img/tree-OPD_svg-tex}    \includegraphics[width=0.64\textwidth]{img/tree-OLOP_svg-tex}
%    \includegraphics[width=0.64\textwidth]{img/tree-KL-OLOP_svg-tex}
%
%    \caption{The look-ahead trees expanded by different planning algorithms for a budget $n=10^3$, shown down to depth 6. The width of edges is proportional to the nodes visit count $T_a(M)$.}
%    \label{fig:trees}
%\end{figure}

### 2019-09-14 paper/poster/poster.tex

\bibliography{../references.bib}

@@

We consider the optimal control of an MDP $\cM = (\cS, A, R, T, \gamma)$ with bounded rewards $R\in[0,1]$
\begin{itemize}[nolistsep]
    \item[\incarrow] $R$ and $T$ are \hlr{unknown}
    \item[\incarrow] Access to a \hlb{generative model} $s' \sim \probability{s'|s,a}$ and $r\sim\probability{r|s, a}$
    \item[\incarrow] \hlr{Fixed-budget} setting: the generative model is \hlr{costly}, can only be queried $n$ times
\end{itemize}
}

@@

%Split the budget $n$ into $M$ sequences of length $L$.
%Let $M$ be the largest integer such that $M \log M/(2 \log 1/\gamma) \leq n$\;
%Let $L = \log M / (2 \log 1/\gamma)$\;
\begin{enumerate}
    \item Sample $M$ sequences of actions of fixed length $L$
    \begin{center}
\includesvg[width=0.8\textwidth]{img/olop-explain.svg}
\end{center}
    \item Use the \hlb{return structure} to \hlb{generalise} to unseen sequences
    \begin{center}
\includesvg[width=0.8\textwidth]{img/olop-explain-2.svg}
\end{center}
\item Be \hlb{Optimistic in the Face of Uncertainty}\\
    \incarrow~ in \hlr{observed} and \hlr{future} rewards
\end{enumerate}

@@

\hlr{Overly pessimistic}, especially in the low-budget regime.

\begin{equation}
         U^{\mu}_a(m) = \hat{\mu}_a(m) + \sqrt{\frac{2 \log M}{T_a(m)}}
    \end{equation}
    \begin{equation}
    \label{eq:Ua}
        U_a(m) \eqdef \sum_{t = 1}^h \gamma^t U^{\mu}_{a_{1:t}}(m) + \frac{\gamma^{h+1}}{1-\gamma}
    \end{equation}
    \begin{equation}
    \label{eq:Ba}
        B_a(m) \eqdef \inf_{1 \leq t \leq L} U_{a_{1:t}}(m)
    \end{equation}
    Intuitive explanation: 
    \begin{itemize}
        \item Unintended behaviour happens when $\hler{U^{\mu}_a(m) > 1}, \forall a$. 
        \item Then the sequence $(U_{a_{1:t}}(m))_t$ is non-decreasing
        \item Then \hlr{$B_a(m) = U_{a_{1:1}}(m)$}
    \end{itemize}

\begin{center}
\includesvg[width=0.8\textwidth]{img/olop-explain-3.svg}    
\end{center}

@@

    \includesvg[width=0.33\textwidth]{../img/tree-ODP}    \includesvg[width=0.33\textwidth]{../img/tree-OLOP}
    \includesvg[width=0.33\textwidth]{../img/tree-KL-OLOP}
    \begin{center}
        \textbf{Expanded trees} for a budget $n=10^3$
    \end{center}
}

>>>>

\bibliography{../biblio.bib}

@@

We consider the optimal control of an MDP $\cM = (\cS, A, R, T, \gamma)$ with bounded rewards $R\in[0,1]$
\begin{itemize}[nolistsep]
    \item[\incarrow] $R$ and $T$ are \hlr{unknown}
    \item[\incarrow] Access to a \hlb{generative model} $s' \sim \probability{s'|s,a}$ and $r\sim\probability{r|s, a}$
    \item[\incarrow] \hlr{Fixed-budget} setting: the generative model is \hlr{costly}, can only be queried $n$ times
\end{itemize}
\texttt{UCT}: \hlr{doubly-exponential} $\xrightarrow{}$ \OPD: \hlg{polynomial}, \hlr{deterministic}
}

@@

%Split the budget $n$ into $M$ sequences of length $L$.
%Let $M$ be the largest integer such that $M \log M/(2 \log 1/\gamma) \leq n$\;
%Let $L = \log M / (2 \log 1/\gamma)$\;
\begin{enumerate}
    \item Sample $M$ sequences of actions of fixed length $L$
    \begin{center}
\includesvg[width=0.75\textwidth]{img/olop-explain.svg}
\end{center}
    \item Use the \hlb{return structure} to \hlb{generalise} to unseen sequences
    \begin{center}
\includesvg[width=0.75\textwidth]{img/olop-explain-2.svg}
\end{center}
\item Be \hlb{Optimistic in the Face of Uncertainty}\\
    \incarrow~ in \hlr{observed} and \hlr{future} rewards
\end{enumerate}

@@

\hlr{Overly optimistic}, especially in the low-budget regime.

    \begin{equation}
         \underbrace{U^{\mu}_a(m)}_{\text{Upper bound}} \eqdef \underbrace{\hat{\mu}_a(m)}_{\text{Empirical mean}} + \underbrace{\sqrt{\frac{2 \log M}{T_a(m)}}}_{\text{Confidence interval}}
    \end{equation}
    \begin{equation}
    \label{eq:Ua}
        U_a(m) \eqdef \sum_{t=1}^h \underbrace{\gamma^t U^{\mu}_{a_{1:t}}(m)}_{\text{Past rewards}} + \underbrace{\frac{\gamma^{h+1}}{1-\gamma}}_{\text{Future rewards}}
    \end{equation}
    \begin{equation}
    \label{eq:Ba}
        B_a(m) \eqdef \inf_{1 \leq t \leq L} U_{a_{1:t}}(m)
    \end{equation}
    Intuitive explanation: 
    \begin{itemize}[noitemsep]
        \item Unintended behaviour happens when $\hler{U^{\mu}_a(m) > 1}, \forall a$. 
        \item Then the sequence $(U_{a_{1:t}}(m))_t$ is non-decreasing
        \item Then \hlr{$B_a(m) = U_{a_{1:1}}(m)$}
    \end{itemize}

\begin{center}
\includesvg[width=0.75\textwidth]{img/olop-explain-3.svg}    
\end{center}

@@

    \includesvg[width=0.33\textwidth]{../img/tree_OPD}    \includesvg[width=0.33\textwidth]{../img/tree_OLOP}
    \includesvg[width=0.33\textwidth]{../img/tree_KL-OLOP}
    \begin{center}
        \textbf{Expanded trees} for a budget $n=10^3$
    \end{center}
}
